Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-03-16 paperDistance/paper.tex
tweaks

\author[0000-0001-5725-9329]{Lauren Anderson}
\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}

@@

We would like to infer the distance to stars using the measured (noisy) parallax from Gaia. Gaia gives us the likelihood of the parallax of a star, given some unknown distance and uncertainties (assumed gaussian)
\begin{equation}
\label{eq:gaialike}
P(\varpi |d, \sigma^2_{\varpi}, \textbf{I})
\end{equation}

where $d$ is distance to the star, $\varpi$ is the observed parallax, and $\sigma^2_{\varpi}$ is the uncertainty in that measurement, and I are other various prior assumptions (mostly about the noise properties?). This is the probability of the data given the model distance, but we want to flip it around and infer the distance to the star given the measured parallax and it's uncertainty. We do this using probability theory/bayes rule, and turn the likelihood of the data into the posterior probability of the distance

\begin{equation}
\label{eq:bayes}
P(d_{true}|\varpi, \sigma^2_{\varpi}, \textbf{I}) = \frac{P(\varpi |d_{true} \sigma^2_{\varpi}, \textbf{I}) P(d)}{\textbf{Z}},
\end{equation}

@@

True values:
  C - true color,
  M - true absolute magnitude
  Y = [10.^0.2(M+2), C]
data:
    (J-K)_c
    \varpi
    J_c
    y = [\varpi10^0.2J_c, (J-K)_c]

  prior assumptions:
    \sigma_J, sigma_K, sigma_\varpi, Greene map

Our true, latent values are C, the true J-K color, and M the true absolute magnitude, which we encapsulate in the true vector $\vec{Y}$. The observed data is the dust corrected J-K color $(J-K)_n$, the Gaia parallax $\varpi_n$, and the dust corrected J band magnitude $J_c$, which we encapsulate in the data vector.  Our prior assumptions $I$ are the uncertainties in each of these measurements, as well as the Greene dust map

\begin{equation}
\label{eq:data}
\begin{aligned}
\mathrm{true vector} \;\;\;\; &\vec{Y} = [10^{0.2\,M + 2}, C] \\
\mathrm{data vector} \;\;\;\; &\vec{y} = [\varpi\,10^{0.2J_{c,n}}, (J- K)_{c,n}] \;  n \in \mathrm{stars} \\
\mathrm{with \; absolute \; magnitude} \;\;\;\; &M_{J,n} = J_n - DM_n - Q_JE(B-V)_n \\
& DM_n = -5\log_{10}\frac{1}{\varpi} \\
\mathrm{and \; color} \;\;\;\; &(J - K)_n = J_n - K_n - Q_{JK}E(B-V)_n \\
\mathrm{assumptions} \;\;\;\; &I = \left\{\left\{\sigma^2_n\right\}_{n=1}^N, \mathrm{assumptions}\right\}
\end{aligned}
\end{equation}

where $E(B-V)$ is the output of the Schlafly dust model, and $Q$ is the correction factor for that band assuming some reddening law.

Likelihood function:
  P(y|Y) = N(y|Y,I)
Working with these variables, we then have a likelihood for the data, both the Gaia parallaxes and 2mass photometry of

\begin{equation}
\label{eq:fulllike}
P(\vec{y}|\vec{Y}) = N(\vec{y}|\vec{Y}, I)
\end{equation}

To generate our prior, we fit the data in the CMD space using XD which deconvolves the data with its noise, which are heterogeneous measurement uncertainties. It generates the denoised, underlying distribution from which the noisey data were drawn from. It represents the distribution as a mixture of gaussians which maximizes the marginalized likelihood of the noisy data, marginalized over the true, latent variables.

XD generates the gaussian parameters

\begin{equation}
\label{eq:xdmml}
\begin{aligned}

\alpha =& \left\{\mathrm{amp}_k, \mu_k, V_k\right\}_{k=1}^K \; k \in \mathrm{Gaussians} \\
       =& \argmax_{\alpha} \Pi_n P(\vec{y}_n|\alpha) \\
       =& \argmax_{\alpha} \Pi_n \int P(\vec{y}_n|\vec{Y}_n,I)P(\vec{Y}|\alpha)d\vec{Y}
\end{aligned}
\end{equation}

\begin{equation}
\label{eq:posterior}
P(Y|y, \alpha, I) = 1/Z p(y| Y,I)p(Y|\alpha, I)
\end{equation}

\begin{equation}
\label{eq:parallaxPost}
P(\vec{Y}_1|\vec{y}, \alpha, I) => p(\varpi_{true}|y, \alpha, I) \, f(\varpi_{true})
\[
f(\varpi_{true}) = \begin{cases}
              1: -2 < log_{10} \varpi < 2\\
              0: otherwise
              \end{cases}
\]
\end{equation}

\end{aligned}
\end{equation}

@@

This work has made use of data from the European Space Agency (\acronym{ESA})
mission Gaia\footnote{\url{http://www.cosmos.esa.int/gaia}}, processed by the Gaia
Data Processing and Analysis Consortium (\acronym{DPAC}\footnote{\url{http://www.cosmos.esa.int/web/gaia/dpac/consortium}}). Funding for the
\acronym{DPAC} has been provided by national institutions, in particular the
institutions participating in the Gaia Multilateral Agreement.

>>>>

\author{Lauren Anderson}
\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}

@@

We would like to infer the distance to stars using the measured (noisy) parallax from Gaia. Gaia gives us the likelihood of the parallax of a star, given some unknown distance and uncertainties (assumed gaussian)
\begin{equation}
\label{eq:likelihood}
P(\varpi |d, \sigma^2_{\varpi}, \textbf{I})
\end{equation}

where $d$ is distance to the star, $\varpi$ is the observed parallax, and $\sigma^2_{\varpi}$ is the uncertainty in that measurement, and I are other various prior assumptions (mostly about the noise properties?). This is the probability of the data given the model, or distance, but we want to flip it around and infer the distance to the star given the measured parallax and it's uncertainty. We do this using probability theory/bayes rule, and turn the likelihood of the data into the posterior probablity of the distance

\begin{equation}
\label{eq:bayes}
P(d|\varpi, \sigma^2_{\varpi}, \textbf{I}) = \frac{P(\varpi |d, \sigma^2_{\varpi}, \textbf{I}) P(d)}{\textbf{Z}},
\end{equation}

@@

\begin{equation}
\label{eq:data}
\begin{aligned}
\mathrm{data vector} \;\;\;\; &\vec{y} = [10^{0.2M_{J,n}}, (J- K)_n] \;  n \in \mathrm{stars} \\
\mathrm{with \; absolute \; magnitude} \;\;\;\; &M_{J,n} = J_n - DM_n - Q_JE(B-V)_n \\
& DM_n = -5\log_{10}\frac{1}{\varpi} \\
\mathrm{and \; color} \;\;\;\; &(J - K)_n = J_n - K_n - Q_{JK}E(B-V)_n
\end{aligned}
\end{equation}

where $E(B-V)$ is the output of the Schlafly dust model, and $Q$ is the correction factor for that band assuming some reddening law. To generate our prior, we fit the data in the CMD space using XD which deconvolves the data with its noise, which are heterogeneous measurement uncertainties. It generates the denoised, underlying distribution from which the noisey data were drawn from. It represents the distribution as a mixture of gaussians which maximizes the likelihood of the noisy data.

\begin{equation}
\label{eq:xdlikelihood}
P(\vec{y}|\alpha, I)
\end{equation}

with the output mixture of gaussians

\begin{equation}
\label{eq:xd}
\begin{aligned}
\alpha = &\left\{\mathrm{amp}_k, \mu_k, V_k\right\}_{k=1}^K \; k \in \mathrm{Gaussians} \\
I = &\left\{\left\{\sigma^2_n\right\}_{n=1}^N, \mathrm{assumptions}\right\}
\end{aligned}
\end{equation}

@@

This work has made use of data from the European Space Agency (\acronym{ESA})
mission \gaia\footnote{\url{http://www.cosmos.esa.int/gaia}}, processed by the Gaia
Data Processing and Analysis Consortium\footnote{\url{http://www.cosmos.esa.int/web/gaia/dpac/consortium}} (\acronym{DPAC}). Funding for the
\acronym{DPAC} has been provided by national institutions, in particular the
institutions participating in the Gaia Multilateral Agreement.
